---
title: "Multilevel Regression"
author: "Samantha-Jo Caetano"
date: "November 15, 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Getting Started

Let's first load in the tidyverse and lme4 packages.

```{r}
library(tidyverse)
#install.packages("lme4")
library(lme4)
#install.packages("merTools")
library(merTools)
```



```{r}
# ?hsb or help(hsb) to learn a bit more about the data.

hsb_data <- hsb

glimpse(hsb_data)
```

# Exploratory Data Analysis

Let's just get to know the data a bit. We are gonna be predicting math scores (mathach) and we will be using the following variables at some point: school id (schid), socioeconomic status (ses), gender (female), private school (schtype). Let's just get to know each of these variables a bit before going into the modelling.

## Math Scores

```{r}
hist(hsb_data$mathach)
boxplot(hsb_data$mathach)
summary(hsb_data$mathach)

hsb_data %>% summarise(n=n(),
                       mean=mean(mathach),
                       sd=sd(mathach),
                       min = min(mathach),
                       max = max(mathach))

ggplot(data = hsb_data, aes(x=mathach)) + 
  geom_histogram(colour="black", fill="grey", bins=15) +
  theme_minimal() 

ggplot(data = hsb_data, aes(y=mathach)) + 
  geom_boxplot(colour="black", fill="grey") +
  theme_minimal() +
  theme(axis.ticks.x = element_blank(), # removes x-axis ticks and scale
        axis.text.x = element_blank()) 

```

## School ID

```{r}
hsb_data %>% group_by(schid) %>% 
                      summarise(n=n(),
                       mean=mean(mathach),
                       sd=sd(mathach),
                       min = min(mathach),
                       max = max(mathach))


# how many distinct schools are there?

hsb_data %>% group_by(schid) %>% summarise() %>% summarise(n=n())

ggplot(data = hsb_data, aes(y=mathach, x=schid)) + 
  geom_boxplot(colour="black", fill="grey") +
  theme_minimal() 
```

## Socioeconomic Status

```{r}
## Math Scores

hist(hsb_data$ses)
boxplot(hsb_data$ses)
summary(hsb_data$ses)

hsb_data %>% summarise(n=n(),
                       mean=mean(ses),
                       sd=sd(ses),
                       min = min(ses),
                       max = max(ses))

## SES vs Math Scores
plot(hsb_data$ses, hsb_data$mathach)
cor(hsb_data$ses, hsb_data$mathach)

```

## Gender

```{r}
hsb_data %>% group_by(female) %>% 
                      summarise(n=n(),
                       mean=mean(mathach),
                       sd=sd(mathach),
                       min = min(mathach),
                       max = max(mathach))

ggplot(data = hsb_data, aes(x=female)) + 
  geom_bar(colour="black", fill="grey") +
  theme_minimal() 

ggplot(data = hsb_data, aes(y=mathach, x=as.factor(female))) + 
  geom_boxplot(colour="black", fill="grey") +
  theme_minimal() 
```

## Private School

```{r}
hsb_data %>% group_by(schtype) %>% 
                      summarise(n=n(),
                       mean=mean(mathach),
                       median = median(mathach),
                       sd=sd(mathach),
                       min = min(mathach),
                       max = max(mathach))

ggplot(data = hsb_data, aes(x=schtype)) + 
  geom_bar(colour="black", fill="grey") +
  theme_minimal() 

ggplot(data = hsb_data, aes(y=mathach, x=as.factor(schtype))) + 
  geom_boxplot(colour="black", fill="grey") +
  theme_minimal() 
```


\newpage

# Linear Regression

## Random Intercept Models

### With no Level I predictors

Let's try to look at student math scores based solely on school id as a level 2 variable.

```{r}
model1 <- lmer(mathach ~ 1 + (1|schid), REML=FALSE, data=hsb)

# REML - Restricted Maximum Likelihood Estimation
summary(model1)
confint(model1)
```

Recall, there is an alternative way to write the model.

```{r}
alt_model1 <- lmer(mathach ~ (1|schid), REML=FALSE, data=hsb)
```

We can even make a prediction. What is the predicted math score of a student who went to school 1224?

```{r}
predict(model1, tibble(schid="1224"))
predict(alt_model1, tibble(schid="1224"))
```

Let's compare that with some of our EDA.

```{r}
hsb_data %>% summarise(n=n(), mean=mean(mathach))

hsb_data %>% group_by(schid) %>% 
                      summarise(n=n(),
                       mean=mean(mathach))
```

### With Level I Predictors

Let's predict math scores with female and SES as our level I predictors still using a random intercept model, with school as the level II variable.

```{r}
model2 <- lmer(mathach ~ 1+female+ses+(1|schid), REML=FALSE, data=hsb)
alt_model2 <- lmer(mathach ~ female+ses+(1|schid), REML=FALSE, data=hsb)

# REML - Restricted Maximum Likelihood Estimation
summary(model2)
confint(model2)
```


What is the predicted math score of a female student with SES of 0.4 who went to school 6816?

```{r}
predict(model2, tibble(schid="6816", female=1, ses=0.4))
```


## Random Slope Model

Let's assume that we want to model the SES slope based on the school cluster. 

```{r}
model3 <- lmer(mathach ~ female+ses+(ses|schid), REML=FALSE, data=hsb)

# REML - Restricted Maximum Likelihood Estimation
summary(model3)
confint(model3)
```

# Logsitic Regression

Let's build similar models, but assume our outcome is instead "scored in top 50th percentile" (binary). Let's create this new outcome variable here:

```{r}
median(hsb_data$mathach)

hsb_data <- hsb_data %>% mutate(top50math = case_when(mathach>=median(hsb_data$mathach)~1,
                                          mathach<median(hsb_data$mathach)~0))
```


Okay, now that we have our outcome variable, let's build a random intercept model, with just school id.

```{r}
logmodel1 <- glmer(top50math ~ (1|schid), data=hsb_data, family="binomial")
summary(logmodel1)
confint(logmodel1)
```

Again, we can even make a prediction. What is the predicted probability of scoring in top 50 percentile of a student who went to school 1224?

```{r}
predict(logmodel1, tibble(schid="1224"), type="response")

### OR

exp(predict(logmodel1, tibble(schid="1224")))/(1+exp(predict(logmodel1, tibble(schid="1224"))))
```


Okay, now let's add in those level I variables (female and ses).

```{r}
logmodel2 <- glmer(top50math ~ female+ses+(1|schid), data=hsb_data, family="binomial")
summary(logmodel2)
confint(logmodel2)
```

Again, we can make a prediction here. What is the predicted probability of scoring in top 50 percentile of a male student with SES 1.2 who went to school 1224?

```{r}
predict(logmodel2, tibble(schid="1224", female=0, ses=1.2), type="response")

### OR

exp(predict(logmodel2,  tibble(schid="1224", female=0, ses=1.2)))/(1+exp(predict(logmodel2,  tibble(schid="1224", female=0, ses=1.2))))
```


