---
title: "Multilevel Regression"
author: "Samantha-Jo Caetano"
date: "November 15, 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Getting Started

Let's first load in the tidyverse and lme4 packages.

```{r}
library(tidyverse)
#install.packages("lme4")
library(lme4)
#install.packages("merTools")
library(merTools) # This is where the data is
```



```{r}
# ?hsb or help(hsb) to learn a bit more about the data.

hsb_data <- hsb

glimpse(hsb_data)
```

# Exploratory Data Analysis

Let's just get to know the data a bit. We are gonna be predicting math scores (mathach) and we will be using the following variables at some point: school id (schid) and socioeconomic status (ses). Let's just get to know each of these variables a bit before going into the modelling.

## Math Scores

Create a histogram, boxplot and some summary measures of math scores:

```{r}

```

## School ID

Create some plots and some summary tables to give us a sense of school id and it's association with math scores:

```{r}


# how many distinct schools are there? (pollev.com/sta) - adapt the above code to count the number of schools.

 
```

## Socioeconomic Status

Create some plots and some summary tables to give us a sense of ses variable and it's association with math scores.

```{r}

# what is the correlation between ses and math scores for all observations (i.e., not "accounting" for school)?


```

## School ID, SES & Math Scores

Let's look at the relationship between ses and math scores, for the different schools. Let's create some scatterplots.

```{r}

```


Okay, let's look at the mean math scores, mean ses and correlation of math scores and SES across different schools. Create a summary table.

```{r}

```

Let's plot the above three to see what they look like diagnostically. You could also create another summary table of the summary table.

```{r}


```


There is probably more depth to go into. Feel free to explore on your own later.

## Extra Exploratory Data Analysis

### Gender

```{r}
hsb_data %>% group_by(female) %>% 
                      summarise(n=n(),
                       mean=mean(mathach),
                       sd=sd(mathach),
                       min = min(mathach),
                       max = max(mathach))

ggplot(data = hsb_data, aes(x=female)) + 
  geom_bar(colour="black", fill="grey") +
  theme_minimal() 

ggplot(data = hsb_data, aes(y=mathach, x=as.factor(female))) + 
  geom_boxplot(colour="black", fill="grey") +
  theme_minimal() 
```

### Private School

```{r}
hsb_data %>% group_by(schtype) %>% 
                      summarise(n=n(),
                       mean=mean(mathach),
                       median = median(mathach),
                       sd=sd(mathach),
                       min = min(mathach),
                       max = max(mathach))

ggplot(data = hsb_data, aes(x=schtype)) + 
  geom_bar(colour="black", fill="grey") +
  theme_minimal() 

ggplot(data = hsb_data, aes(y=mathach, x=as.factor(schtype))) + 
  geom_boxplot(colour="black", fill="grey") +
  theme_minimal() 
```


\newpage

# Linear Regression

## Random Intercept Models

### With no Level I predictors

Let's try to look at student math scores based solely on school id as a level 2 variable.

So we are estimating: $$Y_{ij} = \beta_0 + v_{0j} + \epsilon_{ij} $$
where $Y$ is math scores. 

This is like a baseline model.

```{r}
model1 <- lmer(mathach ~ 1 + (1|schid), REML=FALSE, data=hsb_data)

# REML - Restricted Maximum Likelihood Estimation

summary(model1)
```



Below will give you the confidence intervals of parameters estimated by the model.
```{r}
confint(model1)
```


Recall, there is an alternative way to write the model.

```{r}
alt_model1 <- lmer(mathach ~ (1|schid), REML=FALSE, data=hsb_data)
```

We can even make a prediction. What is the predicted math score of a student who went to school 1224?

```{r}

```


### With Level I Predictors

Let's predict math scores with SES as a level I predictor still using a random intercept model, with school as the level II variable. So we are estimating: $$Y_{ij} = \beta_0 + \beta_1 X_{ij} + v_{0j} + \epsilon_{ij} $$
where $y$ is math scores and $x$ is the SES.

```{r}
model2 <- lmer(mathach ~ 1+ses+(1|schid), REML=FALSE, data=hsb)
alt_model2 <- lmer(mathach ~ ses+(1|schid), REML=FALSE, data=hsb)

summary(model2)
```




What is the predicted math score of a female student with SES of 0.4 who went to school 6816? Go to pollev.com/sta to answer this.

```{r}

```


## Random Slope Model

Let's assume that we want to model the SES slope based on the school cluster. 

$$Y_{ij} = \beta_0 + \beta_1 X_{ij} + v_{0j} + w_jX_{ij} + \epsilon_{ij} $$
where $y$ is math scores and $x$ is the SES.

```{r}
model3 <- lmer(mathach ~ ses+(ses|schid), REML=FALSE, data=hsb)

summary(model3)
```





# Logsitic Regression

Let's build similar models, but assume our outcome is instead "scored in top 50th percentile" (binary). Let's create this new outcome variable here:

```{r}
median(hsb_data$mathach)

hsb_data <- hsb_data %>% 
  mutate(top50math = 
           case_when(mathach>=median(hsb_data$mathach)~1,
                     mathach<median(hsb_data$mathach)~0))
```


Okay, now that we have our outcome variable, let's build a random intercept model, with just school id.

```{r}
logmodel1 <- glmer(top50math ~ (1|schid), data=hsb_data, family="binomial")
summary(logmodel1)
```

Again, we can even make a prediction. What is the predicted probability of scoring in top 50 percentile of a student who went to school 1224? Go to pollev.com/sta to answer this.

```{r}

```


Okay, now let's add in those level I variables (female and ses).

```{r}
logmodel2 <- glmer(top50math ~ female+ses+(1|schid), data=hsb_data, family="binomial")
summary(logmodel2)
```

Again, we can make a prediction here. What is the predicted probability of scoring in top 50 percentile of a male student with SES 1.2 who went to school 1224?

```{r}

```




